Attributions for Poverty in Post-Socialist Countries

### 2014-03-27 article2013.Rmd
eastern europe viittaukset vaihdettu korrektimpiin. samoin lahteet lisatty. jatin kuvioihin viela Central Eastern Europe ryhmittelyn. eikohan se siina kelpaa. terveiset Moskovan junasta Lahden ja Kouvolan valilta. Just puski hytti tayteen venalaisia votkakassit heiluen. ihmettelevat hamletia ylahyllylla. jatkan slaideista huomenna.

---
title: "Attributions for Poverty in Post-Socialist Countries"
author: [Markus Kainu & Mikko Niemelä]
affiliation: Finland
date: March 3, 2014
header-includes:
    - \usepackage{fancyhdr}
    - \pagestyle{fancy}
    - \fancyhead[CO,CE]{}
    - \fancyfoot[CO,CE]{}
    - \fancyfoot[LE,RO]{\thepage}
abstract: Utilising The Life in Transition Survey (LiTS) conducted in 2010, the purpose of the paper is to examine public perceptions of the causes of poverty in Europe and Western Asia with special emphasis on post-socialist countries. The study applies multilevel techniques in order to analyse the role of individual and country-level explanatory factors as determinants of poverty attributions. As country-level determinants, paper analyses the changes in country's economic performance during the period of global financial crises, income inequality and the state of democracy. As individual-level determinants, the paper tests to what extent risk factors related to financial hardship are associated with perceptions of the causes of poverty.
keywords: poverty
output:
  pdf_document:
    toc: true
    number_section: true
---

@@

The focus of this study is on the public perceptions of the causes of poverty in post-socialist countries. Since the early 1990s, the development of post-socialist welfare state has received abundant scholarly attention (e.g. Cerami and Vanhuysse 2009; Inglot 2008; Haggard and Kauffman 2008; Cook 2007; Aidukaite 2004; Deacon 1992). As the decades under socialist regime have had an undeniable influence on societies, these studies have emphasized the role of institutional differences within region as well as pre-socialist historical origins of the welfare states. Now after 20 years of transition, as there has not been a visible convergence towards a common "post-socialist model", the attention has shifted even more on various institutional and historical explanations (Szikra and Tomka 2009; Tomka 2006; Cerami and Vanhuysse 2009; Inglot 2008).

The purpose of this article is to examine cross-national differences in attributions for poverty in Europe and Western Asia with special emphasis on post-socialist countries. Prior findings of the comparative studies on public perceptions of the causes of poverty have emphasized clear differences between Western and Eastern European countries (Gallie and Paugam 2002; Kallio and Niemelä 2014; Lepianka 2007; van Oorschot and Halman 2000). East Europeans are more inclined to blame either the individual behavior or the flaws of the economic system of country than their Western counterparts. Hence, it seems that unlike in Western Europe traditionally opposite attitudinal dimensions, i.e. structural and individualistic explanations of poverty coincide with each other in East Europeans' perceptions. However, there is shortage of comparative studies on attributions for poverty in particular focusing on the East Central European countries or on the whole post-socialist space. There is therefore a need for a more detailed and up-to-date analysis of the public perceptions of the causes of poverty in post-socialist countries.

The novelty of this study is that it covers wide range of countries from Europe and Western Asia. Thus, compared to previous studies, it provides more valid evidence on the differences in perceptions between the East and the West and within post-socialist regime as well as on the importance of the post-socialist legacy. In addition, the study applies multilevel techniques in order to analyze the role of individual and contextual level explanatory factors as determinants of poverty attributions. The article explores to what extent specific risks related to household financial circumstances at the individual-level and the state of democracy, income inequality and changes in national economy at the contextual level are associated with different types of lay explanations of poverty. Finally, extensive country-selection of our data provides a novel opportunity to examine the effects of the financial crisis both at the individual and contextual level and their association with attributions for poverty. At the individual-level, the study focuses not only on the individuals’ current financial situation. Instead, the article provides an empirical example of how past experiences, current financial position as well as future expectations concerning individual’s economic situation are associated with perceptions of the causes of poverty.

@@

Prior empirical findings on attributions for poverty in Europe have shown that there are some differences between the West and the East of Europe. Western Europeans are more inclined to support social explanations for poverty than their Eastern counterparts. On the other hand, individual blame type of explanation is generally more popular in the East than in the West of Europe (Kallio and Niemelä 2014; Kreidl 2000; Lepianka 2007; Lepianka et al. 2010; van Oorschot and Halman 2000). Earlier results suggest, however, that there are variations both within and across groups of countries. Interestingly enough, the variation is highest within the East Central European countries (Kallio and Niemelä 2014). Also Lepianka et al. (2010) come to the conclusion that further investigation is required due to rather large variation across countries. They showed that individual fate type of explanation is supported relatively less frequently in the East of Europe than in Western European countries. Regarding the variation within the Eastern European countries, they found that in some countries individual blame is strongly endorsed: especially the citizens of Czech Republic attribute poverty in internal terms (see also Lepianka 2007). On the other hand, there are countries like Lithuania, Poland, Romania and Croatia in which social blame type of explanation is strongly supported. Finally, social fate type of explanation – living in poverty is explained as simply being part of modern progress – is endorsed especially in Russia.

@@

Besides the so called good government, national-level economic conditions have been proved to be associated with welfare attitudes. This is particularly interesting in these times when the Western world is meeting the increasing economic turbulence. In regard to East-West differences as well as variation within post-socialist countries, countries examined in this study have faced global economic crisis in distinct ways. While many Western and Eastern European countries have witnessed economic decline, the economic performance in Western Asia has been strong. Hence, by analyzing the role of changes in economic performance during the period of global financial crisis, the research design of this article provides empirical evidence on the effects of crisis to public perceptions of the causes of poverty. Prior studies have shown that increased financial strain is associated with stronger support for state responsibility for economic provision (Blekesaune 2007). In regard to explanations of poverty, Gallie and Paugam (2002, pp. 21-24) found that in the majority of European countries, there was a marked decline of support for the individual blame explanation between 1989 and 1993 as economic conditions deteriorated, followed  by an increase between 1993 and 2001 when economic conditions improved. This result is confirmed by Kallio and Niemelä (2014). They found that the larger the economic growth in a country is, more likely people are to blame the poor themselves. Thus, we can assume that changes in country’s economic performance are associated with attributions for poverty regardless of country grouping.

@@

Earlier studies have shown that attributions for poverty differ in general between the West and the East of Europe but the variation between East Central European countries is relatively large. There is, however, a shortage of comparative studies which have focused particularly on differences within post-socialist countries. In addition, most of the prior research is based on attitude surveys from the 1990s and most of them have restricted country selection. Therefore, as previous studies (Kallio and Niemelä 2014; Lepianka et al. 2010) have suggested, there is a need for detailed and more up-to-date analysis which would focus on the attributions for poverty within the post-socialist space. Hence, the general purpose of the study is to analyze public perceptions of the causes of poverty in Europe and Central Asia with special emphasis on post-socialist countries.

The study utilizes the data from the second wave of Life in Transition Survey[^***2] (LiTS2), conducted by European Bank of Development and Reconstruction (EBRD) and  World Bank in 2010. The analyses cover 22 post-socialist countries and 4 Western European countries. The survey was collected using a two-stage clustered stratified sampling procedure to select the households that were included in the sample (for more detailed description of sampling and data collection, see ERBD 2011). The survey had a special focus on how people’s lives were affected by the global economic crises and its aftermath.

@@

In regard to differences between the East and the West of Europe, results show that there are some differences. Yet, there are also substantial variation within Western Europe, which correspond earlier findings on cross-national differences in Western Europe (Lepianka et al. 2010; Kallio and Niemelä 2014). Clearest differences between the West and the East is the stronger support for social fate in Western Europe. In Germany and Sweden external reasons for poverty receive the majority of support. On the other hand, attributions for poverty in France and Italy seems to be quite similar with post-socialist countries - blame types of explanations are endorsed more likely than fate types of explanations. Finally, the figures of Great Britain are the example of selective welfare policy dominated liberal regime in which the individual blame type of explanations is strongly supported (see also Albrekt Larsen 2006).

@@

Analyses regarding the determinants of poverty explanations starts by bivariate associations between the dependent variables and three contextual variables (Figures 3, 4 and 5). Correlation coefficients within the post-socialist countries are shown on top of each regression line. Figure 3 shows the association between attributions for poverty and the change in the size of economy during the period of global financial crisis (2007-2010). In regard to post-socialist countries there is substantial variation between countries in economic change. The trend is very clear: the less country has suffered from the financial crisis the less there is support to social blame type of explanation (-.62) and more to individual blame (.71). Thus, the results are in line with previous studies which suggest positive correlation between the economic growth and support for the individualistic perception of the causes of poverty (Gallie and Paugam 2002; Kallio and Niemelä 2014). In regard to East-West differences, we can see that the variation in economic situation is much smaller in Western European countries, and the association between economic change and attributions for poverty is opposite to the post-socialist countries. However, this is greatly due to economic decline in Great Britain. Hence, without Great Britain we can not see any significant association between economic change and poverty perceptions in Western Europe.

@@

To sum up, the results indicate that in general social blame type of explanation is clearly the most popular explanation in post-socialist countries. Social blame is followed by individual blame, social fate and individual fate types of explanations. Hence, the results are in line with our assumptions and prior research (Habibov 2011; Lepianka et al. 2010) regarding the general support for different explanations of poverty in the post-socialist space. In regard to differences between the East and the West of Europe, results show that Western Europeans emphasize more social reasons for poverty.

@@

As a methodological contribution, findings showed interesting differences between the East and the West regarding the effects of contextual variables. Even though the selection of Western European countries in this study included countries from different welfare regimes, the differences between countries in economic change, democracy and living conditions are quite modest compared especially to CIS countries. Consequently our results showed more modest and in some cases opposite associations between dependent and contextual variables among Western European countries than among post-socialist group.

@@

DEACON 1992 MARKUS OBS

@@

ERBD 2011 MARKUS OBS

@@

GELMAN & HILL 2007 MARKUS OBS

@@

Schumpeter, J. A. (1942) Capitalism, Socialism and Democracy (New York, Routledge).

SNIJDERS & BOSKER 2011 MARKUS OBS

@@

SZIKRA & TOMKA 2009 MARKUS OBS

@@

TOMKA 2006 MARKUS OBS

@@

library(texreg)
texreg(list(soc.4,ind.4,socVsInc),
       float.pos="H",
       #use.packages=TRUE,
       #booktabs=TRUE,
       #dcolumn=TRUE,
       scriptsize=TRUE

>>>>

---
title: "Attributions for Poverty in Post-Socialist Countries"
author: [Markus Kainu & Mikko Niemelä]
affiliation: Finland
date: March 27, 2014
header-includes:
    - \usepackage{fancyhdr}
    - \pagestyle{fancy}
    - \fancyhead[CO,CE]{}
    - \fancyfoot[CO,CE]{}
    - \fancyfoot[LE,RO]{\thepage}
abstract: Utilising The Life in Transition Survey (LiTS) conducted in 2010, the purpose of the paper is to examine public perceptions of the causes of poverty in Europe and Western Asia with special emphasis on post-socialist countries. The study applies multilevel techniques in order to analyse the role of individual and country-level explanatory factors as determinants of poverty attributions. As country-level determinants, paper analyses the changes in country's economic performance during the period of global financial crises, income inequality and the state of democracy. As individual-level determinants, the paper tests to what extent risk factors related to financial hardship are associated with perceptions of the causes of poverty.
keywords: poverty
output:
  pdf_document:
    toc: true
    number_section: true
---

@@

The focus of this study is on the public perceptions of the causes of poverty in post-socialist countries. Since the early 1990s, the development of post-socialist welfare state has received abundant scholarly attention (e.g. Cerami and Vanhuysse 2009; Inglot 2008; Haggard and Kauffman 2008; Cook 2007; Aidukaite 2004; Deacon 1992). As the decades under socialist regime have had an undeniable influence on societies, these studies have emphasized the role of institutional differences within region as well as pre-socialist historical origins of the welfare states. Now after 25 years of transition, as there has not been a visible convergence towards a common "post-socialist model", the attention has shifted even more on various institutional and historical explanations (Szikra and Tomka 2009; Tomka 2006; Cerami and Vanhuysse 2009; Inglot 2008).

The purpose of this article is to examine cross-national differences in attributions for poverty in Europe and Western Asia with special emphasis on post-socialist countries. Prior findings of the comparative studies on public perceptions of the causes of poverty have emphasized clear differences between Western and post-socialist European countries (Gallie and Paugam 2002; Kallio and Niemelä 2014; Lepianka 2007; van Oorschot and Halman 2000). People in post-potsocialist Europe are more inclined to blame either the individual behavior or the flaws of the economic system of country than their Western counterparts. Hence, it seems that unlike in Western Europe traditionally opposite attitudinal dimensions, i.e. structural and individualistic explanations of poverty coincide with each other in post-socialist Europeans' perceptions. However, there is shortage of comparative studies on attributions for poverty in particular focusing on the post-socialist European countries or on the whole post-socialist space. There is therefore a need for a more detailed and up-to-date analysis of the public perceptions of the causes of poverty in post-socialist countries.

The novelty of this study is that it covers wide range of countries from Europe and Western Asia. Thus, compared to previous studies, it provides more valid evidence on the differences in perceptions between the post-socialist countries and the West and within post-socialist space. In addition, the study applies multilevel techniques in order to analyze the role of individual and contextual level explanatory factors as determinants of poverty attributions. The article explores to what extent specific risks related to household financial circumstances at the individual-level and the state of democracy, income inequality and changes in national economy at the contextual level are associated with different types of lay explanations of poverty. Finally, extensive country-selection of our data provides a novel opportunity to examine the effects of the financial crisis both at the individual and contextual level and their association with attributions for poverty. At the individual-level, the study focuses not only on the individuals’ current financial situation. Instead, the article provides an empirical example of how past experiences, current financial position as well as future expectations concerning individual’s economic situation are associated with perceptions of the causes of poverty.

@@

Prior empirical findings on attributions for poverty in Europe have shown that there are some differences between the Western and the post-socialist Europe. Western Europeans are more inclined to support social explanations for poverty than their post-socialist Europeans counterparts. On the other hand, individual blame type of explanation is generally more popular in the post-socialist Europe than in the West of Europe (Kallio and Niemelä 2014; Kreidl 2000; Lepianka 2007; Lepianka et al. 2010; van Oorschot and Halman 2000). Earlier results suggest, however, that there are variations both within and across groups of countries. Interestingly enough, the variation is highest within the post-socialist European countries (Kallio and Niemelä 2014). Also Lepianka et al. (2010) come to the conclusion that further investigation is required due to rather large variation across countries. They showed that individual fate type of explanation is supported relatively less frequently in the post-socialist Europe than in Western European countries. Regarding the variation within the post-socialist European countries, they found that in some countries individual blame is strongly endorsed: especially the citizens of Czech Republic attribute poverty in internal terms (see also Lepianka 2007). On the other hand, there are countries like Lithuania, Poland, Romania and Croatia in which social blame type of explanation is strongly supported. Finally, social fate type of explanation – living in poverty is explained as simply being part of modern progress – is endorsed especially in Russia.

@@

Besides the so called good government, national-level economic conditions have been proved to be associated with welfare attitudes. This is particularly interesting in these times when the Western world is meeting the increasing economic turbulence. In regard to differences between post-socialist countries and Western Europe as well as variation within post-socialist countries, countries examined in this study have faced global economic crisis in distinct ways. While many Western and post-socialist European countries have witnessed economic decline, the economic performance in Western Asia has been strong. Hence, by analyzing the role of changes in economic performance during the period of global financial crisis, the research design of this article provides empirical evidence on the effects of crisis to public perceptions of the causes of poverty. Prior studies have shown that increased financial strain is associated with stronger support for state responsibility for economic provision (Blekesaune 2007). In regard to explanations of poverty, Gallie and Paugam (2002, pp. 21-24) found that in the majority of European countries, there was a marked decline of support for the individual blame explanation between 1989 and 1993 as economic conditions deteriorated, followed  by an increase between 1993 and 2001 when economic conditions improved. This result is confirmed by Kallio and Niemelä (2014). They found that the larger the economic growth in a country is, more likely people are to blame the poor themselves. Thus, we can assume that changes in country’s economic performance are associated with attributions for poverty regardless of country grouping.

@@

Earlier studies have shown that attributions for poverty differ in general between the Western and the post-socialist Europe but the variation between post-socialist European countries is relatively large. There is, however, a shortage of comparative studies which have focused particularly on differences within post-socialist countries. In addition, most of the prior research is based on attitude surveys from the 1990s and most of them have restricted country selection. Therefore, as previous studies (Kallio and Niemelä 2014; Lepianka et al. 2010) have suggested, there is a need for detailed and more up-to-date analysis which would focus on the attributions for poverty within the post-socialist space. Hence, the general purpose of the study is to analyze public perceptions of the causes of poverty in Europe and Central Asia with special emphasis on post-socialist countries.

The study utilizes the data from the second wave of Life in Transition Survey[^***2] (LiTS2), conducted by European Bank of Development and Reconstruction (EBRD) and  World Bank in 2010. The analyses cover 22 post-socialist countries and 5 Western European countries. The survey was collected using a two-stage clustered stratified sampling procedure to select the households that were included in the sample (for more detailed description of sampling and data collection, see EBRD 2011). The survey had a special focus on how people’s lives were affected by the global economic crises and its aftermath.

@@

In regard to differences between the Western and the post-socialist Europe, results show that there are some differences. Yet, there are also substantial variation within Western Europe, which correspond earlier findings on cross-national differences in Western Europe (Lepianka et al. 2010; Kallio and Niemelä 2014). Clearest differences between the Western and the post-socialist Europe is the stronger support for social fate in Western Europe. In Germany and Sweden external reasons for poverty receive the majority of support. On the other hand, attributions for poverty in France and Italy seems to be quite similar with post-socialist countries - blame types of explanations are endorsed more likely than fate types of explanations. Finally, the figures of Great Britain are the example of selective welfare policy dominated liberal regime in which the individual blame type of explanations is strongly supported (see also Albrekt Larsen 2006).

@@

Analyses regarding the determinants of poverty explanations starts by bivariate associations between the dependent variables and three contextual variables (Figures 3, 4 and 5). Correlation coefficients within the post-socialist countries are shown on top of each regression line. Figure 3 shows the association between attributions for poverty and the change in the size of economy during the period of global financial crisis (2007-2010). In regard to post-socialist countries there is substantial variation between countries in economic change. The trend is very clear: the less country has suffered from the financial crisis the less there is support to social blame type of explanation (-.62) and more to individual blame (.71). Thus, the results are in line with previous studies which suggest positive correlation between the economic growth and support for the individualistic perception of the causes of poverty (Gallie and Paugam 2002; Kallio and Niemelä 2014). In regard to differences between Western Europe and the post-socialist countries, we can see that the variation in economic situation is much smaller in Western European countries, and the association between economic change and attributions for poverty is opposite to the post-socialist countries. However, this is greatly due to economic decline in Great Britain. Hence, without Great Britain we can not see any significant association between economic change and poverty perceptions in Western Europe.

@@

To sum up, the results indicate that in general social blame type of explanation is clearly the most popular explanation in post-socialist countries. Social blame is followed by individual blame, social fate and individual fate types of explanations. Hence, the results are in line with our assumptions and prior research (Habibov 2011; Lepianka et al. 2010) regarding the general support for different explanations of poverty in the post-socialist space. In regard to differences between the Western and the post-socialist Europe, results show that Western Europeans emphasize more social reasons for poverty.

@@

As a methodological contribution, findings showed interesting differences between the Western Europe and the post-socialist countries regarding the effects of contextual variables. Even though the selection of Western European countries in this study included countries from different welfare regimes, the differences between countries in economic change, democracy and living conditions are quite modest compared especially to CIS countries. Consequently our results showed more modest and in some cases opposite associations between dependent and contextual variables among Western European countries than among post-socialist group.

@@

Deacon, B., Castle-Kanerova, M., Manning, N., Millard, F., Orosz, E., Szalai, J. & Vidinova, A. (1992) 'The New Eastern Europe: social policy past, present and future', Sage Publications London/Newbury Park.

@@

EBRD (2011) 'Transition Report 2011. Crisis and Transition: The People's Perspective', European Bank of Reconstruction and Development.

@@

Gelman, A. & Hill, J. (2007) 'Data Analysis Using Regression And Multilevel/Hierarchical Models', Cambridge University Press.

@@

Schumpeter, J. A. (1942) 'Capitalism, Socialism and Democracy'. (New York, Routledge).

Snijders, T. A. B. & Bosker, R. (2011) 'Multilevel Analysis: An Introduction to Basic and Advanced Multilevel Modeling', Second Edition. Sage Publications Ltd.

@@

Szikra, D. & Tomka B. (2009) 'Social Policy in East Central Europe: Major Trends in the Twentieht Century', In Cerami, A, ja Vanhuysse, P. (2009) 'Post-Communist Welfare Pathways: Theorizing Social Policy Transformations in Central and Eastern Europe', Palgrave Macmillan,

@@

Tomka, B. (2006) 'East Central Europe and the European Social Policy Model: A Long-Term View'. East European Quarterly 40, nro 2: 135.

@@

library(texreg)
texreg(list(soc.4,ind.4,socVsInc),
       float.pos="H",
       #use.packages=TRUE,
       #booktabs=TRUE,
       #dcolumn=TRUE,
       scriptsize=TRUE,
        custom.model.names = model.names.multi.2,
        custom.coef.names = coef.names.multi,
       include.pvalues=TRUE,
        caption="Logistic multilevel random intercept model comparing the full models for both dependent variable with model predicting social blame against individual blame", 
        caption.above=TRUE)
```
